---
title: 'Self-organized criticality in repository-mediated projects'
author: "J. J. Merelo"
date: "29 de agosto de 2016"
output: pdf_document
bibliography: biblio.bib
abstract: |
   In this paper we will examine if a particular source-contro system repository, devoted to the collaborative writing of a software paper, is in a self-organized critical state, which can be measured by the existence of a scale-free structure, long-distance correlations and pink noise when analyzing the size of changes and its time series.
   Our intention is to prove that, although with different characteristics, any repository independently of the number of collaborators and its real nature, self-organizes, which implies that it is the nature of the interactions, and not the object of the interaction, which takes the project to a critical state. At the same time, we will examine the differences with software projects. 
---

```{r setup, include=FALSE}
library(ggplot2)
library("ggfortify")
library(dplyr)
library(TTR)
#use 
file <- '/home/jmerelo/proyectos/literaturame/data/papers/lines_2015_books_ALL_tex.csv'
# for testing here
lines <- read.csv(file) # File should be established from an R script
lines$SMA5 <- SMA(lines$Lines.changed,n=5)
lines$SMA10 <- SMA(lines$Lines.changed,n=10)
```

#Introduction

The existence of a self-organized critical state [@bak1988self] in software repositories has been well established [@wu2007empirical] [@gorshenev2004punctuated] [@Merelo2016] and attributed to an stigmergy process [@robles05] in which collaborators interact through the code itself and other communication media, such as the Slack chat application, task assignment systems or mailing lists. The case for this critical state is supported by mainly the non-existence of a particular scale in the size of changes [@wu2007empirical] [@gorshenev2004punctuated], but in some cases they also exhibit long-range correlations and a *pink noise* in the power spectral density, with *noise* or variations with respect to the *normal* frequency changing in a way that is inversely proportional to it, higher frequency changes getting a smaller spectral density. 

After examining and establishing the existence of this state in the software repository for the Moose library, in this report we are going to work on a repository for a paper our research group has been working for a long time. Since our research group supports open science, we develop in open software repositories, and this one can be found in https://github.com/geneura-papers/2015_books . This particular paper has been chosen since it has been a work in progress for more than one a year, and in fact in this precise moment it is in the process of being reviewed to be submitted again to a journal. 

Coming up next, we will examine the methodology followed to extract information from the repository and how we have processed it. [Results](#res) will be presented in the next section, and eventually we will expose our [conclusions](#conc).

# Methodology {#methodology}
To extract information of changes in this file, changes have been circumscribed to the paper itself: two files including the paper and response to reviewers, which are the ones that more particularly relate to the projects. Other files are considered artifacts (for instance, images) and are not taken into account. The repository is analyzed with a Perl script and a sequence of lines changed in each commit is generated. Since changes include insertion and deletion of lines within those files, the biggest of these values is taken; in particular, this means that the addition of all changes will not be equal to the sum of the sizes of all files. A change in two lines will appear in a diff as "2 insertions, 2 deletions", adding up to 0; that is why we consider the bigger of these two values.


#Results {#res}

This is a summary of the characteristics of commit sizes.
```{r summary}
summary(lines)
```

The timeline of the commit sizes is represented next, in different ways, with logarithmic or decimal *y* scale. The serrated characteristics is the same, as well as the big changes in scale. *x* axis is simply the temporal sequence of commits, while the *y* axis is the absolute size. 

```{r timeline, echo=FALSE}
lines$x = as.numeric(row.names(lines))
gg.lines = ggplot(lines)
gg.lines+geom_point(aes(x=x,y=Lines.changed))
gg.lines+geom_line(aes(x=x,y=Lines.changed,group=1))
gg.lines+geom_point(aes(x=x,y=Lines.changed))+scale_y_log10()
gg.lines+geom_line(aes(x=x,y=Lines.changed,group=1))+scale_y_log10()
```


Probably it will be a bit more visible if we do a order 5 and then an order 10 smoothing
```{r smoothie, echo=FALSE}
gg.lines+geom_line(aes(x=x,y=SMA5,color='SMA5'))+geom_line(aes(x=x,y=SMA10,color='SMA10'))+scale_y_log10()
```

The change in scale might mean that it is distributed using a Pareto distribution. Next we represent the number of changes of a particular size in a log-log scale.

```{r linecount, echo=FALSE}
by.lines <- group_by(lines,Lines.changed)
lines.count <- summarize(by.lines, count=n())
sizes.fit <- lm(log(1+lines.count$Lines.changed) ~ log(lines.count$count))
ggplot(lines.count, aes(x=Lines.changed, y=count))+geom_point()+scale_x_log10()+scale_y_log10()+stat_smooth()
```

This distribution also appears as a Zipf distribution, with the commit sizes ranked in descending order and plotted with a logarighmic *y* axis. I can be linearly fit to a log-log distribution with coefficients `r sizes.fit$coefficients`

```{r powerlaw, echo=FALSE}
ggplot(data=lines, aes(lines$Lines.changed)) + geom_histogram(bins=100)+scale_x_log10()
sorted.lines <- data.frame(x=1:length(lines$Lines.changed),Lines.changed=as.numeric(lines[order(-lines$Lines.changed),]$Lines.changed))
ggplot()+geom_point(data=sorted.lines,aes(x=x,y=Lines.changed))
ggplot()+geom_point(data=sorted.lines,aes(x=x,y=Lines.changed))+scale_y_log10()
sorted.lines.no0 <- sorted.lines[sorted.lines$Lines.changed>0,]
zipf.fit <- lm(log(sorted.lines.no0$Lines.changed) ~ sorted.lines.no0$x)
```

The Zipf exponent for this graph is `r zipf.fit$coefficients`

Finally, these scale distributions hints at the possibility of long-scale correlations. The partial autocorrelation is plotted in the next figure.

```{r autocorrelation, echo=FALSE}
autoplot(pacf(lines$Lines.changed, plot=FALSE) )
```

These possible long range autocorrelations point also to a state of self-organized criticality. One of the characteristics of this state is the presence of *pink* noise, as measured by the power spectral density, shown below

```{r spectrum, echo=FALSE}
autoplot(spectrum(lines$Lines.changed, plot=FALSE) )
```

##Conclusions {#conc}

This files has been generated automatically from the data file, so it is difficult to draw a conclusion sight unseen. From what I have seen, there are in many cases long-distance correlations centered around 20 commits of distance, and they show mainly in repositories with a collaborative activity, but they might show up too in some others developed mainly by a single person. You can draw your own conclusions on your own repos by running the Perl script in <http://github.com/JJ/literaturame>

# References