---
title: "Complex graphs in repositories"
author: "JJ Merelo"
date: "13 de febrero de 2017"
output: html_document
---

##Abstract

>There is no single *natural* way of analyzing the graph created by the collaborative work of a team of software developers as reflected in the log of that repository. In this paper we will analyze several graphs created by this work, considering author or files the nodes in what is actually a bipartite network. We will try to find out which network better represents the complex nature of the technosocial system.

```{r setup, echo=FALSE}
library("igraph")

# from http://stackoverflow.com/questions/5643002/r-selectively-display-nodes-when-plotting-an-igraph
delete.isolates <- function(graph) {
  bad.vs<-V(graph)[degree(graph) == 0]
  delete.vertices(graph, bad.vs)
}
beehive.author.net <- read_graph(file = "../data/sna/author-beehive.net", format = 'pajek')
beehive.author.net.core <- delete.isolates(beehive.author.net)
beehive.files.net <- read_graph(file = "../data/sna/files-beehive.net", format = 'pajek')
beehive.files.net.core <- delete.isolates(beehive.files.net)
beehive.commit.net <- read_graph(file = "../data/sna/commit-beehive.net", format = 'pajek')
beehive.commit.net.core <- delete.isolates(beehive.commit.net)

```

# Introduction

Beehive is a Go event and agent system available with a free license from https://github.com/muesli/beehive.

The repository can be represented as a bipartite graph with authors and files, with non-directed edges connecting authors with those files they have worked on. There are several possible ways of turning this bipartite graph into a one-party graph: create a graph of files joined by one edge if they have one developer in common, or joining authors by one edge if they have worked on at least one file. These are represented next

```{r plot, echo=FALSE}
plot(beehive.author.net.core)
plot(beehive.files.net.core)
```

However, the file graph is not entirely satisfactory since, in fact, the main developer might have touched all files, and, second, files completely unrelated might be joined since they are created or modified sequentially by the same developer. 


```{r commit, echo=FALSE}
plot(beehive.commit.net.core)
```

